Waggle run distributions quantify collective foraging in honey bee colonies

### 2022-03-15 analysis/manuscript/paper.Rmd
updated ms with latest edits

**Figure of honeybee foraging**

@@

Honeybees provide an ideal study system in which to evaluate collective foraging thanks to the famous “waggle dance”, which conveys information about location of a resource as a vector of distance (duration of the waggle run) and direction (angle of the dance relative to the sun) [@Frisch1993].Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Resource “quality” is provided through the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]). Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A].

**End Box 1**

@@

For 17 out of 20 study hives, a model of collective foraging provided a better explanation of the data than one of individual foraging (Fig. 3a). In all but one of these cases, the collective model had a good fit (using a Kolmogorov-Smirnov statistic, see methods) to the empirical waggle run durations, whereas for 9 of these sites, the distribution predicted by the scout model was significantly different to the observed data (Fig. 3b). The sites shown in Figs 3c-d are representative examples showing the model fits, showing the overall quality of the model fit to the data.

```{r, fig.cap="Figure 3. A. Location of study hives in Southern England coloured by which model, collective (black circles) or individual (red triangles) provided the best explanation of the data as indicated by lowest AIC score. Sites STU and ZSL (C & D) are marked. Inset plot shows the location of the sites within the UK; shaded area in the main plot indicates Greater London. B. Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. P value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data. The red dashed line marks the significance threshold of 0.05. Values greater than 0.05 indicate no statistically significant difference is observed between the model and the data, indicating the model provides a good fit to the data. C. Model fits to waggle run durations collected from the ZSL hive, for which the model did not outperform the individual model, along with histogram showing distribution. D. Model fits to waggle run durations collected from the STU hive along with histogram showing distribution. Model fits show the compliment cumulative frequency distribution, giving the probability of sampling a value greater than or equal to x."}
knitr::include_graphics("../results/figures/sites_model_plot.png")
```

@@

In the agri-rural environments one combination of land use types explained ~52% of the variation in the proportion of scouts (beta regression: $R^2 = 0.52$, $\phi$ = 3.7, p < 0.05, Fig. 4. A). This principal component correlates positively with arable land (29% of land coverage; Table 1) and negatively with built-up areas (17% of land coverage) ; it also correlates negatively with non-agricultrual unmanaged green space and water , but together these represent less than 3% of land-use (Fig. 4. B). These land-use types maintain a significant correlation with PC1 over the jackknifed PLS (Fig. 4. B) indicating the results are robust. As arable land increases whilst built-up areas decrease, the proportion of trips that are driven by individual search increases.

@@

A similar effect was observed in the urban environment, where a single PC explained ~59% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.59, $\phi$ = 6.2, p < 0.05, Fig. 5. A). This PC correlates positively with parks, allotments and cemeteries (8% of land cover; Table 2), whilst correlating negatively with sparse residential lands (35% of land cover) and railways but these represent only 1% of land cover. (Fig. 5. B). However, we have low confidence in this finding because the jackknifed PLS identifies CAD as a highly influential site, the removal of which causes sparse residential land to have a positive correlation with PC1 (Supplementary Material _._).

@@

Our results within the agri-rural sites are in broad agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006; Price2019]. These results suggest that, even though collective foraging is not always beneficial, that in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. These benefits of dancing appear to outweigh the costs associated with dancing and could explain why this remarkable behaviour is present in modern-day bees. However, the results for our urban sites do not produce a clear biological rational and present an all together more varied picture. An alternative hypothesis could be that the the dance is influenced by resource stability more than abundance and that longer term effects may contribute to the benefits of the dance [@Schurch2014]. Whilst the limiting number of sites used in this study handicaps our ability to progress this debate here, as a proof of concept, the methods derived here provide a new toolkit for evaluating factors influencing waggle dance recruitment

>>>>

@@

*Honeybees provide an ideal study system in which to evaluate collective foraging thanks to the famous “waggle dance”, which conveys information about location of a resource as a vector of distance (duration of the waggle run) and direction (angle of the dance relative to the sun) [@Frisch1993].Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Resource “quality” is provided through the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A].*

[end of figure of waggle dance]

@@

For 17 out of 20 study hives, a model of collective foraging provided a better explanation of the data than one of individual foraging (Fig. 3a). In all but one of these cases, the collective model had a good fit (using a Kolmogorov-Smirnov statistic, see methods) to the empirical waggle run durations, whereas for 9 of these sites, the distribution predicted by the scout model was significantly different to the observed data (Fig. 3b). The sites STU and ZSL shown in Figs 3c-d are representative examples showing the model fits where the collective (STU) and individual (ZSL) models fit best, showing the overall quality of the model fit to the data.

```{r, fig.cap="Figure 3. A. Location of study hives in Southern England coloured by which model, collective (black circles) or individual (red triangles) provided the best explanation of the data as indicated by lowest AIC score. Sites STU and ZSL (C & D) are marked. Inset plot shows the location of the sites within the UK; shaded area in the main plot indicates Greater London. B. Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. P value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data. The red dashed line marks the significance threshold of 0.05. Values greater than 0.05 indicate no statistically significant difference is observed between the model and the data, indicating the model provides a good fit to the data. C. Model fits to waggle run durations collected from the STU hive, for which the collective model (red line) did not outperform the individual model (blue line), along with histogram showing distribution. D. Model fits to waggle run durations collected from the ZSL hive, where the collective model (red line) outperformed the individual model (blue line), along with histogram showing distribution. Model fits show the compliment cumulative frequency distribution, giving the probability of sampling a value greater than or equal to x."}
knitr::include_graphics("../results/figures/sites_model_plot.png")
```

@@

In the agri-rural environments one combination of land use types explained ~52% of the variation in the proportion of scouts (beta regression: $R^2 = 0.52$, $\phi$ = 3.7, p < 0.05, Fig. 4. A). This principal component correlates positively with arable land (29% of land coverage; Table 1) and negatively with built-up areas (17% of land coverage) ; it also correlates negatively with non-agricultrual unmanaged green space and water, but together these represent less than 3% of land-use (Fig. 4. B). These land-use types maintain a significant correlation with PC1 over the jackknifed PLS (Fig. 4. B) indicating the results are robust. As arable land increases whilst built-up areas decrease, the proportion of trips that are driven by individual search increases.

@@

A similar effect was observed in the urban environment, where a single PC explained ~59% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.59, $\phi$ = 6.2, p < 0.05, Fig. 5. A). This PC correlates positively with parks, allotments and cemeteries (8% of land cover; Table 2), whilst correlating negatively with sparse residential lands (35% of land cover) and railways, but these represent only 1% of land cover. (Fig. 5. B). However, we have low confidence in this finding because the jackknifed PLS identifies CAD as a highly influential site, the removal of which causes sparse residential land to have a positive correlation with PC1 (Supplementary Material _._).

@@

Our results within the agri-rural sites are in broad agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006; @Price2019]. These results suggest that, even though collective foraging is not always beneficial, that in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. These benefits of dancing appear to outweigh the costs associated with dancing and could explain why this remarkable behaviour is present in modern-day bees. However, the results for our urban sites do not produce a clear biological rational and present an all together more varied picture. An alternative hypothesis could be that the the dance is influenced by resource stability more than abundance and that longer term effects may contribute to the benefits of the dance [@Schurch2014]. Whilst the limiting number of sites used in this study handicaps our ability to progress this debate here, as a proof of concept, the methods derived here provide a new toolkit for evaluating factors influencing waggle dance recruitment
